A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-06 rigid_bodies.tex
Advancements in NVT integration schemes

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,amsfont]{revtex4-1}

@@

\raggedbottom

@@

In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles. In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$. An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$). Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta. Of course, at least $2N$ integrals of motion are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.

@@

Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\[
\mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\]

@@

Therefore, a symplectic integrator can be devised using the same splitting scheme as described in Sec.~\ref{sec:symplectic}. The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\vt P = \sum_{i=1}^N \vt p_i$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\vt G = \sum_{i=1}^N(\vt p_i t - m_i \vt r_i)$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

@@

Understanding of the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space, which is the cornerstone for deterministic molecular dynamics at different ensembles. Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation} \label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric. This can be shown by using the chain rule to express the time-derivative of an arbitrary scalar function $f(\vt x)$, which yields
\[
\dot{f} = \tr{\left(\diff{f}{\vt x}\right)}\dot{\vt x} = \tr{\left(\diff{f}{\vt x}\right)} \boldsymbol{ \mathcal B } \diff{H}{\vt x}.
\]

If $f = H$, the equation above turns into a quadratic form, which is identically null if $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{\mathcal B}$, thus resulting in $\dot H = 0$.

In what follows, index $i$ ranges for all $N$ particles and index $j$ ranges for all $M$ thermostats, unless otherwise stated.

\begin{equation}
\label{eq:H_nvt}
\begin{split}
H = U(\vt r, \vt q) + \sum_{i=1}^N \left( \frac{\tr{\vt p}_i {\vt p}_i}{2 m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8} \right) + \\
+ \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j
\end{split}
\end{equation}

The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \qquad \text{and}\\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat. In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat. The result is
\[
{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
\]

Each other thermostat in the chain acts on the previous one, and is acted upon by the following one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
\[
{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
\]

Finally, the last thermostat just acts on 
\[
{\dot p}_{\eta_M} = -\diff{H}{\eta_M} + p_{\eta_{M-1}} \diff{H}{p_{\eta_{M-1}}}.
\]

For a three-dimensional system composed of non-collinear rigid bodies, the phase-space compressibility is given by\cite{Tuckerman2010}
\[
\kappa = -6N \diff{H}{p_{\eta_1}} - \sum_{j=2}^M \diff{H}{p_{\eta_j}} = -6N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
\]

Thus, the metric determinant factor of the extended phase space is then given by $\sqrt{g} = e^{6N \eta_1 - \sum_{j=2}^M \eta_j}$.\cite{Tuckerman_1999, Tuckerman2001}

Finally, the Nos\'{e}-Hoover chain thermostat applied to a system of $N$ rigid bodies becomes
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \\
&\dot{\eta}_j = \alpha_j \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}
\end{align}
\end{subequations}
where $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$.
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - 6N k_B T \\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad (j > 1)
\end{align*}

$Q_1 = 6 N k_B T t_d^2$ and $Q_j = k_B T t_d^2$

In these equations, $i$ indexes the number of rigid bodies from $1$ to $N$, $M$ is the number of thermostats in the chain, $T$ is the specified temperature, $k_b$ is Boltzmann's constant, $6N$ is the number of DOF, and $t_d$ is the characteristic time period of the thermostat chain. The conserved quantity in this case is the extended energy $\mathcal{H}^\prime$ given by
\begin{align*}
\mathcal{H}^\prime = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_k} + 6Nk_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
\end{align*}
with $\mathcal{H}$ defined in Eq.~\ref{eq:H_with_Omega}.

@@

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the Liouville operator can be written as
\[
i\!L = i\!L_\text{NHC} + i\!L_\text{NVE},
\]
where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
\begin{equation}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) \\
&+ \sum_{j=1}^M \left[ \alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$. Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$. Considering the analytical solution above, the NHC operator is factorized as
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 \exp\left[\frac{\Delta t}{4} \left(G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right] \\
&\times \prod_{j=1}^M \exp\left(\frac{\Delta t}{2} \alpha_j \diff{}{\eta_j}\right) \\
&\times \prod_{i=1}^N \exp\left[-\frac{\Delta t}{2} \alpha_1 \left(\tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right)\right] \\
&\times \prod_{j=1}^M \exp\left[\frac{\Delta t}{4} \left( G_j - \alpha_{j+1} p_{\eta_j} \right) \diff{}{p_{\eta_j}}\right].\\
\end{split}
\end{equation}

The remaining factors that appear in the second line of the equation above are evaluated according to\cite{Martyna1996}
\begin{equation}
\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
\end{equation}
\begin{equation}
\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 \frac{\Delta t}{2} \right)
\end{equation}
\begin{equation}
\eta_j = \eta_j^0 + \alpha_j^0 \frac{\Delta t}{2} 
\end{equation}

>>>>

%\documentclass[prb,preprint,preprintnumbers,amsmath,amssymb]{revtex4-1}
%\documentclass[prb,twocolumn,preprintnumbers,amsmath,amssymb]{revtex4-1}
\documentclass[aip,jcp,reprint,amsmath,amssymb,raggedbottom]{revtex4-1}

@@

@@

In this section, we present factorization schemes for generating molecular dynamics algorithms in the microcanonical (NVE) and canonical (NVT) ensembles. In both cases, a three-dimensional system is considered as consisting of $N$ interacting, non-collinear rigid bodies occupying a fixed volume $V$. An index $i$ ranging from $1$ to $N$ indicates the quantities related to each body, including its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$). Thus, the phase space of the whole system now contains $3N$ spatial coordinates and $4N$ quaternion components, along with their conjugate momenta. Of course, $2N$ implicit constraints are known to exist, as $\tr{\vt q}_i{\vt q}_i = 1$ and $\tr{\vt q}_i{\vt \pi}_i = 0$ for all $i$.

@@

Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

@@

Therefore, a symplectic integrator can be devised using the same splitting scheme as described in Sec.~\ref{sec:symplectic}. The only difference is that the action of each propagator (boost, free translation, and free uniaxial rotation) must now occur simultaneously for all rigid bodies in the system. If there are no external forces acting on the bodies, then additional conserved quantities are the total linear momentum $\sum_{i=1}^N \vt p_i = \vt P$ and the generator of infinitesimal Galilean boosts\cite{Ray1999, Schwichtenberg2015} $\sum_{i=1}^N(\vt p_i t - m_i \vt r_i) = \boldsymbol{\mathcal G}$. If periodic boundary conditions are employed, as it is usual in molecular dynamics simulations, then there is no rotational symmetry and, consequently, no conservation of the total angular momentum.

@@

Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space. Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation} \label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$. We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} with $M$ thermostats to a system with $N$ rigid bodies. For this, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE} and $L$ is a constant to be determined. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the first thermostat. In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat. The result is
\[
{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
\]

Each additional thermostat will then act on the preceding one, and be acted upon by the succeeding one. Therefore, for $j$ from $2$ to $M-1$, the equations of motion are
\[
{\dot p}_{\eta_j} = -\diff{H}{\eta_j} + p_{\eta_{j-1}} \diff{H}{p_{\eta_{j-1}}} - p_{\eta_j} \diff{H}{p_{\eta_{j+1}}}.
\]

Finally, as the last thermostat will just act on the penultimate one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above. Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'{e}-Hoover chain become
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
&\dot{\eta}_j = \alpha_j \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}
\end{align}
\end{subequations}

In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_B T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
\end{align*}

As recommended in Ref.~\onlinecite{Martyna1992}, the inertial parameters can be calculated from a characteristic time scale $t_d$ of the system by $Q_1 = L k_B T t_d^2$ and, for $j > 1$, $Q_j = k_B T t_d^2$. Analysis of the probability distribution generated by the equations of motion in Eq.~\ref{eq:nhc_system} can be done via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $\vt P e^{\eta_1} = \vt K$ are integrals of motion, while the center-of-mass position $\vt R$ can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, from the $8N$ individual equations implied in Eqs.~\ref{eq:nhc_q} and \ref{eq:nhc_pi}, $2N$ must be eliminated due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the extended phase space is $\sqrt{g} = e^{6N \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if $L = 6N$.

@@

As done in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999} This implies that we can employ the Trotter-Suzuki splitting to devise a strategy for integrating the equations of motion. First, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian components. In this case, the Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ was defined in Eq.~\ref{eq:full_operator}. It follows that the operator $i\!L_\text{NHC}$ is given by
\begin{equation}
\begin{split}
i\!L_\text{NHC} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^M \alpha_j \diff{}{\eta_j} + \sum_{j=1}^M \left[ (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} \right].
\end{split}
\end{equation}

If the last summation was absent from the equation above, then the action of the propagator $e^{t i\!L_\text{NHC}}$ would be the solution of the equations $\dot{\vt p}_i = -\alpha_1 \vt p_i$ and $\dot{\vt \pi}_i = -\alpha_1 \vt \pi_i$ for all $i \in [1,N]$ and $\dot{\vt p}_i = \alpha_j$ for all $j \in [1,M]$, with all other variables remaining constant at their initial values. These equations can be solved altogether in analytical form, which results in
\begin{subequations}
\begin{align}
&\vt \pi_i = \vt \pi_i^0 \exp\left(-\alpha_1^0 t \right), \\
&\vt p_i = \vt p_i^0 \exp\left(-\alpha_1^0 t \right), \quad \text{and}\\
&\eta_j = \eta_j^0 + \alpha_j^0 t,
\end{align}
\end{subequations}
where the superscript $0$ stands for ``initial value''.

As pointed out by Martyna~\textit{et al}.,\cite{Martyna1996} the part of the operator given by $(G_j - \alpha_j p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}$ can be evaluated analytically using the following equation:
\begin{equation}
\label{eq:p_eta}
\begin{split}
&p_{\eta_j} = p_{\eta_j}^0 + \left( G_j^0 - \alpha_{j+1}^0 p_{\eta_j}^0 \right) \phi(\alpha_{j+1}^0 t) t, \\
\end{split}
\end{equation}
where $\phi(x) = (1-e^{-x})/x$. For small absolute values of $x$, this function can be evaluated by truncating its Taylor series form $\phi(x) = \sum_{n=0}^\infty \frac{(-1)^n x^n}{(n+1)!}$. Four terms result in a sufficiently precise value if $\|x\| \leq 10^{-4}$. Considering the analytical solution above, the NHC operator is factorized as
\begin{equation}
\begin{split}
e^{\frac{\Delta t}{2} i\!L_\text{NHC}} &= \prod_{j=M}^1 e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}} \\
&\times e^{\frac{\Delta t}{2} \left[-\alpha_1 \sum\limits_{i=1}^N \left(\tr{\vt p}_i \frac{\partial}{\partial \vt p_i} + \tr{\vt \pi}_i \frac{\partial}{\partial \vt \pi_i}\right) + \sum\limits_{j=1}^M \alpha_j \frac{\partial}{\partial \eta_j} \right]} \\
&\times \prod_{j=1}^M e^{\frac{\Delta t}{4} (G_j - \alpha_{j+1} p_{\eta_j}) \frac{\partial}{\partial p_{\eta_j}}}.\\
\end{split}
\end{equation}
